SPAMDA: A software tool to manage meteorological data for environmental modelling

### 2019-12-17 latex_manuscript/elsevier/3_Journal_of_Environmental_Management/spamda.tex
Initial commit.

\journal{Environmental Modelling \& Software}

@@

		\title{SPAMDA: Software for Pre-processing and Analysis of Meteorological DAta to build datasets}

		\author[UCO]{Antonio M. G\'omez-Orellana}
		\ead{am.gomez@uco.es}

@@

		\address[UCO]{Department of Computer Science and Numerical Analysis, University of Cordoba, C\'ordoba, Spain.}

		\begin{abstract}
			%Meteorological data play an important role for the comprehension of the environment, and they are extensively used to perform environmental learning. Machine Learning (ML) techniques have become a valuable support in many research areas, improving the performance of traditional statistical procedures. However, such techniques require datasets containing information related to the topic under study, which are not always available in an appropriate format. Preparing these datasets implies a lot of time and effort by the researchers. This paper presents a software tool for creating datasets using meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} (NDBC) and the \textit{National Centers for Environmental Prediction} (NCEP)/\textit{National Center for Atmospheric Research} (NCAR) \textit{Reanalysis Project}. The datasets created by the software are ready to be used as inputs for ML techniques in prediction tasks (classification or regression). As the designed software is able to simplify the creation of the datasets and reduce the time needed for this task, it prevents researchers from performing repetitive technical work, allowing them to concentrate on the study of the meteorological aspects of the data. Therefore, researchers can benefit from it in order to achieve a more efficient exploitation and protection of the environment.
			Meteorological data play an important role for the comprehension of the environment, and they are extensively used to perform environmental learning. Machine Learning (ML) techniques, a valuable support in many research areas, require datasets containing information related to the topic under study, which are not always available in an appropriate format, and its preparation implies a lot of time and effort by the researchers. This paper presents a software for creating datasets using meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. As the designed software is able to simplify the creation of the datasets (ready to be used as inputs for ML techniques in prediction tasks) and reduce the time needed for this task, researchers can benefit from it in order to achieve a more efficient exploitation and protection of the environment.
		\end{abstract}

		\begin{keyword}
			Meteorological data\sep Reanalysis data\sep  Pre-processing\sep Creating datasets\sep  Prediction tasks\sep Marine energy
		\end{keyword}

@@

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. climate change, agriculture, energy, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: directional analysis of sea storms \cite{LAFACE201545}, wind power ramp events prediction \cite{DORADOMORENO2017428}, offshore wind energy assessment \cite{DVORAK20101244}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		On the other hand, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. In \cite{CAMARGONOGUEIRA2016361}, a software tool for designing solar water heating systems is developed, which simulates different situations and finds the best technical and economical solution. In \cite{LO2015293}, an integrated simulation tool for the optimum design of bifacial solar panel with reflectors is presented. This tool can also be used to analyse the performance of the solar cells. A framework for data integration of offshore wind farms is implemented in \cite{NGUYEN2013150} in order to facilitate data exchange and improve operation and maintenance practices. In \cite{GUTIERREZ201369}, a tool based on \textit{Fatigue, Aerodynamics, Structures, and Turbulence} (FAST) code for performing design optimisation of offshore wind turbines is presented, which can analyse a massive group of cases by means of its parametric design capability. A design tool for architects based on \textit{Lighting and Thermal} (LT) method is used in \cite{BAKER2013156}, with the purpose of comparing and optimising design solutions in terms of energy and comfort performance. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modeling System} (SMC), to support different operational levels of headland-bay beach in coastal engineering projects, and an interactive web application, called the \textit{Stream Hydrology And Rainfall Knowledge System} (SHARKS), for analysis of hydrologic and meteorological time series data that can be used for providing valuable insights to urban hydrologic processes is presented in \cite{BRENDEL2019}.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \cite{BHATTACHARYA2017157, BREDE201344}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed to transform this wave energy into electricity. WECs are mechanical devices that convert kinetic energy into electrical energy by means of either the vertical oscillation of waves or the linear motion of them. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \cite{CROWLEY2018159, Abdelkhalik2016, 6898109} need a proper prediction of waves, in order to maximise the wave energy extraction. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most important wave parameters in this regard: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$). Additionally, wave predictions are also helpful for designing offshore structures \cite{CHATZIIOANNOU2017126}, operational works in the sea \cite{DALGIC2015211}, providing technical information to the coastal and coral reef planners in developing countries \cite{CALLAGHAN2018123}, etc.

		Currently, and as a support to traditional study procedures, Machine Learning (ML) techniques \cite{Alpaydin:2004:IML:1036287,Bishop:2006:PRM:1162264} are being widely used in numerous research fields related to classification, regression and optimisation tasks \cite{srivastava2017large}, obtaining significant improvements in the performance of the results. ML methodologies can be used not only by experienced data and computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML algorithms. In this way, ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. The problem is that, in order to apply these methods, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		The information to create these datasets can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a merging or matching process have to be carried out by the researchers to manually create the datasets with the needed information. Moreover, depending on the subject and the ML technique to be applied, or even if the researcher considers other factors in order to improve the results or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets for ML techniques, resulting in an extremely tedious task.

@@

		Up to now, and to the best of our knowledge, this is the first software tool addressing the problem previously discussed, combining meteorological data from NDBC and NNRP. The advantages that SPAMDA offers to the researchers will be detailed in Section \ref{sec:SPAMDA}, although some of them are briefly summarised below:
		\begin{itemize}
			\item The generation of datasets becomes a very easy and customizable task, by means of the selection of different input parameters.
			\item It makes the researcher focus on oceanic and atmospheric studies, without having to worry about mechanical tasks.
			\item It provides information about the quality and quantity of the data.
			\item It avoids possible researcher errors in the intermediate steps of the process of creation of the datasets.
			\item It includes different pre-processing tasks, such as normalisation and missing data recovery.
			\item It facilitates data management and well-organised storage of the datasets.
			\item Its modular design allows the implementation of new functional modules for managing meteorological data from others sources for renewable energy research.
			\item It includes an user-friendly GUI, facilitating and greatly simplifying data management, and it is integrated with the Explorer environment of WEKA.
			\item It is multi-platform, and it can be used on any computer with Java regardless of the operating system.
		\end{itemize}

@@

			Note that the data collected by the network of buoys may be incomplete due to diverse circumstances such as the weather conditions in which the buoys have to operate, failures or malfunctioning elements of the buoys, among others. Accordingly, it may be the situation that some of the measurements are completely missing (missing date or instance) or partially missing (some measurements not recorded), by a buoy or by a set of buoys, once in a while or over a period of time. It may be also possible that the measurements have been recorded at a time different from the expected one. This aspect have to be taken into account when creating the datasets. This casuistry is explained in detail in Appendix A.

@@

					\begin{itemize}
						\item \textit{Normalize} \cite{WEKA_Filter_Normalize}: This filter normalises all numeric values of each attribute. The resulting values are by default in the interval [0,1].
						\item \textit{Remove} \cite{WEKA_Filter_Remove}: It removes an attribute or a range of them.
						\item \textit{RemoveByName} \cite{WEKA_Filter_RemoveByName}: It removes attributes based on a regular expression matched against their names.
						\item \textit{ReplaceMissingValues} \cite{WEKA_Filter_ReplaceMissingValues}: For each attribute, all the missing values will be replaced by the average value of the attribute.
						\item \textit{ReplaceMissingWithUserConstant} \cite{WEKA_Filter_ReplaceMissingWithUserConstant}: This filter replaces all the missing values of the attributes with an user-supplied constant value.
					\end{itemize}

				 \item \textit{Instance}: All theses filters can be applied to the instances (hourly measurements of the buoy from NDBC) of the intermediate dataset.
					\begin{itemize}
						\item \textit{RemoveDuplicates} \cite{WEKA_Filter_RemoveDuplicates}: With this filter, all duplicated instances are removed.
						\item \textit{RemoveWithValues} \cite{WEKA_Filter_RemoveWithValues}: This filter removes all the instances that match the attribute and the value supplied by the user.
						\item \textit{SubsetByExpression} \cite{WEKA_Filter_SubsetByExpression}: It removes all the instances which do not match a user-specified expression.
					\end{itemize}

@@

				In order to merge and format the data provided by the two sources of information described in Section \ref{sec:DataSources}, it is necessary to carry out a matching process. The matching procedure is performed using an intermediate or pre-processed dataset, which includes the measurements collected by a buoy from NDBC, and the needed reanalysis data files from NNRP. Note that SPAMDA is able to manage the NetCDF binary format for handling the information stored in the reanalysis files.

@@

		Although the software includes a user manual, this section will describe how the application works in a practical approach. To do so, an example showing how to create a fully processed dataset (final dataset), starting from the raw data, will be described. The objective of this final dataset is to be used with ML algorithms to classify waves in the Gulf of Alaska depending on their height. The data collected to perform this example is:
		\begin{enumerate}
		\item The measurements obtained from 2013 to 2017 by the buoy with ID 46001, placed in the Gulf of Alaska, which are provided by NDBC as annual text files. This data is publicly available at the NDBC website. 
		\item Complementary information collected from reanalysis data containing air temperature, pressure and two components of wind speed measurements (South-North and West-East). This information will be collected from the four closest reanalysis nodes surrounding the geographical location of the buoy. This data is publicly available at the NNRP website and can be downloaded in NetCDF format.
		\end{enumerate}

		After gathering the information described above\footnote{Further instructions for downloading this data can be found in the user manual of the application.}, the researcher can open SPAMDA.

		In Figure \ref{fig:main_view}, the main window is shown. In order to input the reanalysis data which will be used in further steps for creating the final dataset, the researcher has to select the option \textcolor{blue}{\textit{Manage reanalysis data}}.

@@

		Then, the window of Figure \ref{fig:reanalysis} is shown. Here, using the buttons located at the bottom, it is possible to add, delete or consult any data from the different reanalysis files. After the information has been introduced in the application, this window can be closed and the user can go back to the main window to continue entering the information related to the buoy under study.

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.89\textwidth]{figures/FigureManage_reanalysis_data.png}
			\caption{\textit{Manage reanalysis data} window.}\label{fig:reanalysis}
		\end{figure}

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.89\textwidth]{figures/FigureManage_buoys.png}
			\caption{\textit{Buoys} tab.}\label{fig:manage_buoys}
		\end{figure}

		The researcher has to select \textcolor{blue}{\textit{Manage buoys data}} to open the window shown in Figure \ref{fig:manage_buoys}. In this tab, the researcher can consult, modify, add or delete different data related to the buoy. In order to enter such data, click on the \textcolor{blue}{\textit{New}} button, and then the window shown in Figure \ref{fig:new_buoy} pops up.

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.70\textwidth]{figures/FigureNew_buoy.png}
			\caption{\em{Entering a new buoy} window.}\label{fig:new_buoy}
		\end{figure}

@@

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.90\textwidth]{figures/FigureDatasets.png}
			\caption{\textit{Datasets} tab.}\label{fig:show_datasets}
		\end{figure}

		To create the intermediate dataset, the researcher has to double-click on the buoy under study or click on the \textcolor{blue}{\textit{Datasets}} tab to switch to the corresponding view (see Figure \ref{fig:show_datasets}). In this view, the researcher can delete or consult a summary of each intermediate or pre-processed dataset by selecting it from the corresponding list. It can also create new ones. To proceed with the creation of the intermediate dataset, the user clicks on the \textcolor{blue}{\textit{New}} button, and the view shown in Figure \ref{fig:intermediate} appears.

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.70\textwidth]{figures/FigureNew_intermediate_dataset.png}
			\caption{\textit{New intermediate dataset} view.}\label{fig:intermediate}
		\end{figure}

		Here the researcher can select the annual text files to be included in the intermediate dataset. In this case, all the files introduced before, which correspond to the buoy under study, are selected. When the file selection is finished, \textcolor{blue}{\textit{Create}} button has to be clicked in order to introduce the description and the file name of the current intermediate dataset, and then, with the \textcolor{blue}{\textit{Save}} button, the creation process starts, showing the status of the process during it.

		After that, in order to prepare the intermediate dataset, the dataset is selected, and then the button \textcolor{blue}{\textit{Open}} is clicked to jump to the tab \textcolor{blue}{\textit{Pre-process}} (shown in Figure \ref{fig:preprocess_data}). In this tab, relevant statistical information about the selected dataset is shown, and also the content of the dataset can be consulted, providing the researcher the capacity to evaluate the pre-processing being performed.

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.90\textwidth]{figures/FigurePreprocess.png}
			\caption{\textit{Pre-process} tab.}\label{fig:preprocess_data}
		\end{figure}

		Here the researcher can apply (and configure) the necessary filters (explained in Section \ref{sec:Preprocess}) to the selected dataset, and, in the bottom part, the main statistics of the dataset are displayed, which can be used to observe the changes produced when applying a filter. As mentioned earlier, this case study is focused on classifying waves considering their height, so any missing data from wave height ($376$ values) and the remaining attributes are recovered, using the filter \textit{Replace missing values with symmetric $3$ hours mean}. Furthermore, the attributes MWD, DEWP, VIS and TIDE are removed from the dataset by applying the filter \textit{RemoveByName}, since the first two had more than $92$\% of missing data and the last two $100$\%. After finishing the pre-processing of the dataset, the researcher can click on the \textcolor{blue}{\textit{Save}} button, to introduce the description and file name for the current pre-processed dataset.

		At this point, the researcher has registered the buoy in SPAMDA, then entered its raw data and selected the required data for the problem (intermediate dataset). Finally, the data has been pre-processed in order to be ready for its future use in ML algorithms. In order to achieve a more accurate description of the problem under study, a matching process can be carried out to merge the processed data from NDBC with the reanalysis data (also entered previously) from NNRP. The next step is to click on the \textcolor{blue}{\textit{Matching configuration}} tab, to open the view shown in  Figure \ref{fig:matching_conf}.

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.90\textwidth]{figures/FigureMatching_configuration.png}
			\caption{\textit{Matching configuration} tab.}\label{fig:matching_conf}
		\end{figure}

		In this view, the researcher can customise (or load) the parameters of the matching process according to their needs and select the prediction task  (described in Section \ref{sec:matching_conf}) that the final dataset will be used for. In this example, the following parameters were selected:
		\begin{itemize}
			\item Attribute to predict: WVHT.
			\item Reanalysis data: Air, pressure, u-wind and v-wind.
			\item Buoy attributes to be used as inputs: WDIR, WSPD, GST, DPD, APD, PRES, ATMP and WTMP.
			\item Reanalysis nodes to consider: $1$.
			\item Number of final datasets: In this example that option is disabled, because only one reanalysis node is considered.
			\item Prediction task: Classification.
		\end{itemize}

		\begin{figure}[ht!]
			\centering
			\includegraphics[width=0.90\textwidth]{figures/FigureFinal_datasets.png}
			\caption{\textit{Final datasets} tab.}\label{fig:final_dataset}
		\end{figure}

>>>>

\journal{Journal of Environmental Management}

@@

		\title{SPAMDA: A software tool to manage meteorological data for environmental modelling}

		\author[UCO]{Antonio M. G\'omez-Orellana\corref{AMGO}}
		\cortext[AMGO]{Corresponding author.}
		\ead{am.gomez@uco.es}

@@

		\address[UCO]{Department of Computer Science and Numerical Analysis, University of Cordoba, 14071, C\'ordoba, Spain.}

		\begin{abstract}
			Meteorological data play an important role for the comprehension of the environment, and they are extensively used to perform environmental learning. Machine Learning (ML) techniques, a valuable support in many research areas, require datasets containing information related to the topic under study, which are not always available in an appropriate format, and its preparation implies a lot of time and effort by the researchers. This paper presents a software for creating datasets using meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. As the designed software is able to simplify the creation of the datasets (ready to be used as inputs for ML techniques in prediction tasks) and reduce the time needed for this task, researchers can benefit from it in order to achieve a more efficient exploitation and protection of environmental resources.
		\end{abstract}

		\begin{keyword}
			Meteorological data\sep Reanalysis data\sep Pre-processing\sep Creating datasets\sep Environmental prediction\sep Marine energy
		\end{keyword}

@@

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. climate change, agriculture, energy, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: patterns in transmission of air pollution \cite{LU2019109377}, directional analysis of sea storms \cite{LAFACE201545}, wind power ramp events prediction \cite{DORADOMORENO2017428}, offshore wind energy assessment \cite{DVORAK20101244}, analysis of ecological typologies of large marine areas \cite{DELAHOZ201859}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		On the other hand, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. In \cite{CAMARGONOGUEIRA2016361}, a software tool for designing solar water heating systems is developed, which simulates different situations and finds the best technical and economical solution. In \cite{LO2015293}, an integrated simulation tool for the optimum design of bifacial solar panel with reflectors is presented. This tool can also be used to analyse the performance of the solar cells. A framework for data integration of offshore wind farms is implemented in \cite{NGUYEN2013150} in order to facilitate data exchange and improve operation and maintenance practices. In \cite{PERERA2005291}, a modelling tool for water allocation and management is presented. An interactive system for multidisciplinary wind turbine site selection is developed in \cite{RAFIEE2018107}. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modeling System} (SMC), to support different operational levels of headland-bay beach in coastal engineering projects, and an intuitive tool for stormwater runoff evaluation and management is developed in \cite{CHEN201645}.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \cite{BHATTACHARYA2017157, BREDE201344}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed and widely installed to transform this wave energy into electricity, which can be injected into the electric network or supplied to existing seawater desalination plants \cite{FERNANDEZPRIETO2019546}, among others. WECs are mechanical devices that convert kinetic energy into electrical energy by means of either the vertical oscillation of waves or the linear motion of them. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \cite{CROWLEY2018159, Abdelkhalik2016, 6898109} need a proper prediction of waves, in order to maximise the wave energy extraction. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most important wave parameters in this regard: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$). Additionally, wave predictions are also helpful for designing offshore structures \cite{CHATZIIOANNOU2017126}, operational works in the sea \cite{DALGIC2015211}, providing technical information to the coastal and coral reef planners in developing countries \cite{CALLAGHAN2018123}, etc.

		Currently, and as a support to traditional study procedures, Machine Learning (ML) techniques \cite{Alpaydin:2004:IML:1036287,Bishop:2006:PRM:1162264} are being widely used in numerous research fields related to classification, regression and optimisation tasks, obtaining significant improvements in the performance of the results \cite{MARTINSSILVA2019109368, TORREGROSSA20181061}. ML methodologies can be used not only by experienced computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML algorithms. In this way, ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. The problem is that, in order to apply these methods, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		The information to create these datasets can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a data integration process, denominated as the matching process in this document, has to be carried out by the researchers to manually create the datasets with the needed information. Moreover, depending on the subject and the ML technique to be applied, or even if the researcher considers other factors in order to improve the results or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets for ML techniques, resulting in an extremely tedious task.

@@

		Up to now, and to the best of our knowledge, this is the first software tool addressing the problem previously discussed, combining meteorological data from NDBC and NNRP. The advantages that SPAMDA offers to the researchers will be detailed in Section \ref{sec:SPAMDA}, although some of them are briefly summarised below:
		\begin{itemize}
			\item The generation of datasets becomes a very easy and customizable task, by means of the selection of different input parameters.
			\item The created datasets can be easily used by ML tools.
			\item It makes the researcher focus on environmental modelling, without having to worry about mechanical tasks.
			\item It provides information about the quality and quantity of the data.
			\item It avoids possible researcher errors in the intermediate steps of the process of creation of the datasets, such as time and geographical coordinates conversion or missing values handling, among others.
			\item It includes different pre-processing tasks, such as normalisation and missing data recovery.
			\item It facilitates data management and well-organised storage of the datasets.
			\item Its modular design allows the implementation of new functional modules for managing meteorological data from others sources for environmental research.
			\item It includes an user-friendly GUI, facilitating and greatly simplifying data management, and it is integrated with the Explorer environment of WEKA.
			\item It is multi-platform, and it can be used on any computer with Java regardless of the operating system.
		\end{itemize}

@@

			Note that the data collected by the network of buoys may be incomplete due to diverse circumstances such as the weather conditions in which the buoys have to operate, failures or malfunctioning elements of the buoys, among others. Accordingly, it may be the situation that some of the measurements are completely missing (missing date or instance) or partially missing (some measurements not recorded), by a buoy or by a set of buoys, once in a while or over a period of time. It may be also possible that the measurements have been recorded at a time different from the expected one. These aspects have to be taken into account when creating the datasets. This casuistry is explained in detail in Appendix A \ref{app:AppendixA}.

@@

					\begin{itemize}
						\item \textit{Normalize}: This filter normalises all numeric values of each attribute. The resulting values are by default in the interval [0,1].
						\item \textit{Remove}: It removes an attribute or a range of them.
						\item \textit{RemoveByName}: It removes attributes based on a regular expression matched against their names.
						\item \textit{ReplaceMissingValues}: For each attribute, all the missing values will be replaced by the average value of the attribute.
						\item \textit{ReplaceMissingWithUserConstant}: This filter replaces all the missing values of the attributes with an user-supplied constant value.
					\end{itemize}

				 \item \textit{Instance}: All theses filters can be applied to the instances (hourly measurements of the buoy from NDBC) of the intermediate dataset.
					\begin{itemize}
						\item \textit{RemoveDuplicates}: With this filter, all duplicated instances are removed.
						\item \textit{RemoveWithValues}: This filter removes all the instances that match the attribute and the value supplied by the user.
						\item \textit{SubsetByExpression}: It removes all the instances which do not match a user-specified expression.
					\end{itemize}

@@

				In order to use the data provided by the two sources of information described in Section \ref{sec:DataSources}, it is necessary to carry out a data integration process, denominated as the matching process in this document, to merge and format such data. The matching procedure is performed using an intermediate or pre-processed dataset, which includes the measurements collected by a buoy from NDBC, and the needed reanalysis data files from NNRP. Note that SPAMDA is able to manage the NetCDF binary format for handling the information stored in the reanalysis files.

@@

		This section will describe how the application works in a practical approach. To do so, an example showing how to create a fully processed dataset (final dataset), starting from the raw data, will be described. The objective of this final dataset is to be used with ML algorithms for environmental modelling, concretely, to classify waves in the Gulf of Alaska depending on their height. In this case, the problem will be tackled as a multi-class classification problem, given that any continuous variable can be discretised in different classes. Such waves modelling can be applied with different purposes, energy flux range forecasting, missing buoy data reconstruction, or extreme significant wave heights detection, among others.

		\subsection{Obtaining the final dataset}\label{sec:ObtainingFinalDataset}

			The data collected to perform this example is:
			\begin{enumerate}
			\item The measurements obtained from 2013 to 2017 by the buoy with ID 46001, placed in the Gulf of Alaska, which are provided by NDBC as annual text files. This data is publicly available at the NDBC website. 
			\item Complementary information collected from reanalysis data containing air temperature (air), pressure (pres) and two components of wind speed measurements: South-North (vwind) and West-East (uwind). This information will be collected from the four closest reanalysis nodes surrounding the geographical location of the buoy. This data is publicly available at the NNRP website and can be downloaded in NetCDF format. Concretely, the closest reanalysis nodes downloaded are $57.5$ N $\times$ $147.5$ W, $57.5$ N $\times$ $150$ W, $55$ N $\times$ $147.5$ W and $55$ N $\times$ $150$ W. However, as will be seen later, only the information from the nearest node will be used in the data integration process.
			\end{enumerate}

			After gathering the information described above\footnote{Further instructions for downloading this data can be found in the user manual of the application.}, the researcher can open SPAMDA. In Figure \ref{fig:main_view}, the main window is shown. In order to input the reanalysis data which will be used in further steps for creating the final dataset, the researcher has to select the option \textcolor{blue}{\textit{Manage reanalysis data}}.

@@

			Then, the window of Figure \ref{fig:reanalysis} is shown. Here, using the buttons located at the bottom, it is possible to add, delete or consult any data from the different reanalysis files. Once the information has been introduced in the application, this window can be closed and the user can go back to the main window to continue entering the information related to the buoy under study.

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.89\textwidth]{figures/FigureManage_reanalysis_data.png}
				\caption{\textit{Manage reanalysis data} window: downloaded files containing the four closest reanalysis nodes.}\label{fig:reanalysis}
			\end{figure}

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.89\textwidth]{figures/FigureManage_buoys.png}
				\caption{\textit{Buoys} tab: buoy ID 46001.}\label{fig:manage_buoys}
			\end{figure}

			After that, the researcher has to select \textcolor{blue}{\textit{Manage buoys data}} to open the window shown in Figure \ref{fig:manage_buoys}, where several tabs are available. In \textcolor{blue}{\textit{Buoys}} tab, the researcher can consult, modify, add or delete different data related to the buoy.

			In order to enter such data, click on the \textcolor{blue}{\textit{New}} button, and then the window shown in Figure \ref{fig:new_buoy} pops up.

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.70\textwidth]{figures/FigureNew_buoy.png}
				\caption{\textit{New buoy} window: information of the buoy ID 46001.}\label{fig:new_buoy}
			\end{figure}

@@

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.90\textwidth]{figures/FigureDatasets.png}
				\caption{\textit{Datasets} tab: intermediate datasets of the buoy ID 46001.}\label{fig:show_datasets}
			\end{figure}

			To create the intermediate dataset, the researcher has to double-click on the buoy under study or click on the \textcolor{blue}{\textit{Datasets}} tab (see Figure \ref{fig:manage_buoys}) to switch to the corresponding view (see Figure \ref{fig:show_datasets}). In this view, the researcher can delete or consult a summary of each intermediate or pre-processed dataset by selecting it from the corresponding list. It can also create new ones. To proceed with the creation of the intermediate dataset, the user clicks on the \textcolor{blue}{\textit{New}} button, and the view shown in Figure \ref{fig:intermediate} appears.

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.70\textwidth]{figures/FigureNew_intermediate_dataset.png}
				\caption{\textit{New intermediate dataset} view: creating the intermediate dataset with $5$ annual text files.}\label{fig:intermediate}
			\end{figure}

			Here the researcher can select the annual text files to be included in the intermediate dataset, by clicking on the \textcolor{blue}{\textit{->}} and \textcolor{blue}{\textit{<-}} buttons. In this case, all the files introduced before, which correspond to the buoy under study, are selected. When the file selection is finished, \textcolor{blue}{\textit{Create}} button has to be clicked in order to introduce the description and the file name of the current intermediate dataset, and then, with the \textcolor{blue}{\textit{Save}} button, the creation process starts, showing the status of the process during it. After that, in order to prepare the intermediate dataset, the dataset is selected (see Figure \ref{fig:show_datasets}), and then the button \textcolor{blue}{\textit{Open}} is clicked to jump to the tab \textcolor{blue}{\textit{Pre-process}} (shown in Figure \ref{fig:preprocess_data}).

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.90\textwidth]{figures/FigurePreprocess.png}
				\caption{\textit{Pre-process} tab: preprocessing the created intermediate dataset.}\label{fig:preprocess_data}
			\end{figure}

			In \textcolor{blue}{\textit{Pre-process}} tab, relevant statistical information about the selected dataset is shown, and also the content of the dataset can be consulted, providing the researcher the capacity to evaluate the pre-processing being performed. Here the researcher can apply (and configure) the necessary filters (explained in Section \ref{sec:Preprocess}) to the selected dataset, and, in the bottom part, the main statistics of the dataset are displayed, which can be used to observe the changes produced when applying a filter. As mentioned earlier, this case study is focused on classifying waves considering their height, so any missing data from wave height ($376$ values) and the remaining attributes are recovered, using the filter \textit{Replace missing values with symmetric $3$ hours mean}. Furthermore, the attributes MWD, DEWP, VIS and TIDE are removed from the dataset by applying the filter \textit{RemoveByName}, since the first two had more than $92$\% of missing data and the last two $100$\%. After finishing the pre-processing of the dataset, the researcher can click on the \textcolor{blue}{\textit{Save}} button, to introduce the description and file name for the current pre-processed dataset.

			At this point, the researcher has registered the buoy in SPAMDA, then entered its raw data and selected the required data for the problem (intermediate dataset). Finally, the data has been pre-processed in order to be ready for its future use in ML algorithms. In order to achieve a more accurate description of the problem under study, a matching process can be carried out to merge the processed data from NDBC with the reanalysis data (also entered previously) from NNRP.

			\begin{figure}[ht!]
				\centering
				\includegraphics[width=0.90\textwidth]{figures/FigureMatching_configuration.png}
				\caption{\textit{Matching configuration} tab: parameters for the data integration of the intermediate dataset and the reanalysis files.}\label{fig:matching_conf}
			\end{figure}
